### 2025-04-10 chapters/4-substructure.tex
Final changes to main files for first draft

The principle of \gls{lphd}, discussed in \Def{lphd}, suggests that small changes in hadronic energy flow should be associated with small changes in the associated partonic degrees of freedom, and vice versa.
%
Therefore, we will hope to achieve a way of isolating high-energy, partonic information in a \textit{continuous} way:
%
small changes in the flow of energy within a jet should not drastically affect what we infer about partonic degrees of freedom.
%
Conversely, a jet grooming algorithm is \textit{discontinuous} if small changes in the properties of jet constituents, such as their energies or angular positions, can lead to extensively different groomed results.

@@

    Therefore, the cumulative distribution at LL can be written
    \begin{equation}
    \begin{aligned}
        \label{eq:LL_mass}
        \Sigma^{\rm(LL)}_{m^2}(m^2)
        =
        \sum_{n = 0}^{\infty}\frac{1}{n!}
        \Bigg(
        \frac{\alpha_s}{\pi}
        \int_0^R \frac{\dd\theta}{\theta}
        \int_0^{\frac{1}{2}} \dd z\,[\bar{p}_i(z)]_+^{[1/2]}
        \Bigg[
            \Theta\left(
                \frac{m^2}{Q^2} > z\theta^2
            \right)
            &
            \Theta\left(z > \zcut\right)
            \\
            \notag
            +
            &\,\,\Theta\left(z < \zcut\right)
        \Bigg]
        \Bigg)^n
        \,.
    \end{aligned}
    \end{equation}
    Using the properties of \glspl{plus-fn} as in \Sec{ll-substructure-diagrams}, we can conclude that
    \begin{equation}
    \begin{aligned}
        \Sigma^{\rm(LL)}_{m^2}(m^2)
        =
        \sum_{n = 0}^{\infty}\frac{1}{n!}
        \Bigg(
        \frac{\alpha_s}{\pi}
        \int_0^R \frac{\dd\theta}{\theta}
        \int_0^{\frac{1}{2}} \dd z\,\bar{p}_i(z)
        \Bigg[
            &\Theta\left(
                \frac{m^2}{Q^2} > z\theta^2
            \right)
            \Theta\left(z > \zcut\right)
            \\
            \notag
            &+
            \,\,\Theta\left(z < \zcut\right) - 1
        \Bigg]
        \Bigg)^n
        \,.
    \end{aligned}
    \end{equation}
    Finally, using that \(1 - \Theta(a) = \Theta(-a)\), we can use \(\Theta(z < \zcut) - 1 = -\Theta(z > \zcut)\) to notice that the integrand contains
    \begin{align}
        \le(
        \Theta\left(
            \frac{m^2}{Q^2} > z\theta^2
        \right)
        - 1
        \ri)
        \times
        \Theta\left(z > \zcut\right)
        =
        \Theta\left(
            z\theta^2 > \frac{m^2}{Q^2}
        \right)
        \times
        \Theta\left(z > \zcut\right)
        \,.
    \end{align}

@@

\remark{cc-discontinuity}{
    The discontinuity of the \gls{constant-cutoff} grooming algorithm manifests in LL calculations as a \emph{kinds in the radiators} of substructure observables, and associated discontinuities of LL probability distributions (e.g. as in \Eq{ll-cc-mass})
    %
    Unfortunately, the kink in the radiator presents a complication when attempting to achieve computations beyond LL.
    %
    In particular, the formula of \Eq{multiple-emissions} for the inclusion of multiple-emissions effects on cumulative distributions of additive observables depends on the derivative of the radiator;
    %
    therefore, probability distributions which leverage \Eq{multiple-emissions} depend on the \textit{second derivative} of the radiator.
    %
    Though the radiator turns out to have a well defined derivative, e.g. in \Eq{ll-cc-mass}, it is not second-differentiable, and it is more difficult to account for multiple-emissions effects in \glslink{hard-cutoff-groomer}{hard-cutoff-groomed} jet observables than in ungroomed jet observables.
}

@@

\gls{soft-drop} is a powerful tool which improves upon the strengths of the constant cutoff algorithm.
%
Like the \gls{constant-cutoff} algorithm, \gls{soft-drop} is a tree-based, angular-ordered \gls{declustering} algorithm,%
\footnote{
    As a reminder this means that \gls{soft-drop} acts on a jet with characteristic radius \(R_0\) that is first clustered using an arbitrary algorithm, such as the anti-\(k_t\) algorithm, and then re-clustered with the angular-ordered Cambridge-Aachen (C/A) algorithm.
    %
    As discussed in \Secs{reclustering}{constantcutoff}, this produces a binary tree of clustered particles within the jet, such that particles deeper in the tree tend to be closer in angle.
}
%
and is therefore computationally efficient and amenable to resummed calculations.
%
However, Soft Drop only needs to loop over a subset of the branches of the jet tree, unlike the \gls{constant-cutoff} algorithm, which loops over every branch.
%
Since common sources of contaminating radiation in jets tend to be wide angle, as touched upon in \App{pollution-models}, \gls{soft-drop} can be geared to more effectively remove low-energy pollution from collider events while retaining any low-energy information associated with the process of jet formation and the hard scattering.

@@

We would also like to understand the analytic behavior of the substructure of jets groomed with Soft Drop.
%
Luckily, the \glspl{substructure-diagram} we developed for use with the Constant Cutoff grooming algorithm can be quickly extended for use with Soft Drop.
%
Let us start with the most similar examples, for which \(\gls{beta-sd}=0\).

@@

We begin with a toy example of a \PIRANHA{} groomer in \Sec{constant-cutoff} -- the constant subtraction algorithm, similar to the constant cutoff algorithm of \Sec{constant-cutoff} -- before introducing the three concrete implementations of \PIRANHA{} discussed in this thesis:%
\footnote{
    Code for these three \PIRANHA{} groomers is available on GitHub \cite{piranhagithub}.
    %
    We delegate discussions of \PIRANHA{} grooming in EMD mode to \App{grooming_in_emd_mode}.
}
\begin{itemize}
    \item
    \vocab{\gls{rs}} is a family of nearly-continuous, computationally efficient, tree-based algorithms.
    % motivated by P-AS and P-IVS.
    %
    We develop the concrete implementation of \vocab{\glslink{rsf}{Recursive Subtraction with a Fraction \(f_\text{soft}\) (P-RSF\(_{f}\))}} in \Sec{recursive-subtraction}.
    %
    We discuss how P-RSF\(_f\) is discontinuous in suppressed regions of parameter space, and focus on \PRSF{1/2}, which is the only fully soft-continuous version of P-RSF\(_f\);

@@

\PIRANHA{} groomers may be roughly imagined as optimally transporting hungry piranhas to eat up delicious low-energy pollution.
%
Given an ungroomed event with energy density \(\mathcal{E}_{0}\) in the pseudorapidity-azimuth angle plane and any ansatz \(\rhoC\) for the energy density of low-energy pollution, one particularly succinct \PIRANHA{} algorithm is
%
\sam{haven't introduced EMD yet}
%
\sam{borrow from talks}
\begin{align}
    \mathcal{E}_g\left[\mathcal{E}_0\right]
    =
    \underset{\mathcal{E}'}{\arg \min}\ {\rm EMD}_{\beta, R}
    \left(\mathcal{E}_0, \mathcal{E}' + \rhoC\right)
    \label{eq:simple_piranha_intro}
    ,
\end{align}
where \(\mathcal{E}_g\) is a groomed energy distribution, \(\rho\,\mathcal{C}\) is a free input to the algorithm indicating a model of the energy flow of the contaminating radiation (see \Sec{emd}), and explicit dependence on the EMD metric makes the continuity of \(\mathcal{E}_g\left[\mathcal{E}_0\right]\) manifest.

\gls{apollonius} is a direct, continuous application of \Eq{simple_piranha_intro} with \(\beta = 1\), and for which \(\rhoC\) is uniform in the pseudorapidity-azimuth plane to approximate common sources of \gls{additive-contamination} such as \gls{pileup} and \gls{ue} \cite{Soyez:2018opl,Monk:2018clo,Sjostrand:1987su,Sjostrand:2014zea,Dasgupta:2007wa,Kirchgaesser:2020poq,Moraes:2007rq,CDF:2015txs,Larkoski:2021hee,Baron:2020xoi,Marzani:2017kqd}.
%
\gls{ivs} approximates \gls{apollonius} and preserves its continuity.
%
Finally, the \glslink{rsf}{Recursive Subtraction with a Fraction 1/2 (\PRSF{1/2})} algorithm we introduce in \Sec{rsf} furnishes, to our knowledge, the first example of a soft continuous tree-based jet grooming algorithm;
%
the soft discontinuities of other P-RS algorithms occur only in highly suppressed regions of phase space.

This section is devoted to an introduction of the \PIRANHA{} paradigm.
%
Readers looking for a more detailed discussion of continuity may want to skip to \Sec{softdicont}, which examines the soft discontinuities present in hard-cutoff groomers and \gls{rsf}, the soft continuity of \PRSF{1/2}, and angular discontinuity in tree-based groomers.

@@

@@

We are starting to see why the name \PIRANHA{} is suitable;
%
not only does this algorithm annihilate the soft particles associated with \glslink{pileup}{pileup} and infra-red radiation, but it also can be imagined as putting a fish on each softer branch with a hunger for energy (fraction) given by \zcut, and which will subtract from its assigned branch until either it eats its fill or it completely annihilates its meal.

@@

Up to terms that are power-suppressed in \(C\), \(\zcut\), or both, the fixed-order distribution for the groomed observable away from \(C = 0\) is
%
\begin{subequations}
\label{eq:piranha_nlo}
\begin{align}
    \rho^{\rm(NLO)}_{i,~\varsigma}(C>0)
    &=
    \frac{2\alpha_s C_{R_i}}{\varsigma~\pi}
    \frac{1}{C}
    \left(
        2\tanh^{-1}\left(1  -2 \tilde C - 2\zcut\right)
        + B_i
        +
        \mathcal{O}\left(C,\,\zcut\right)
    \right)
    \\
    &=
    \frac{2\alpha_s C_{R_i}}{\varsigma~\pi}
    \frac{1}{C}
    \left(
        -\log\left(\tilde C + \zcut\right)
        + B_i
        +
        \mathcal{O}\left(C,\,\zcut\right)
    \right)
\end{align}
\end{subequations}
%
where \(\tilde C = C \times (1-\zcut)^2\), \(C_{R_i}\) is the quadratic Casimir of the color representation of the initiating parton, and we have defined the factor \(B_i\) to capture behavior from non-singular components of the splitting functions:
%
\(B_q = -3/4\) for quark initiated jets and \(B_g = -11/12 + n_f/(6 C_A)\) for gluon initiated jets.
\end{example}

@@

\noindent
The plus-function regularized expression of \Eq{plusdist_prsf_nlo} may be derived by replacing \(\overline{p}(z)\) in \Eq{NLO_dist_setup} with its plus-function analog.
%
Alternatively, one can recognize that the full distribution for \(C_1^{(\varsigma)}\) must integrate to one even at \glslink{accuracy}{NLO}, and that additional contributions to the \glslink{accuracy}{NLO} distribution in \Eq{prsf_LO} can only come from the regions of phase space where \(C_1^{(\varsigma)}=0\);
%
this \gls{sum-rule} is enough to uniquely specify the result of \Eq{plusdist_prsf_lo} at \glslink{accuracy}{NLO}.
\end{exercise}

\remark{}{
    Since all \PIRANHA{} algorithms are subtractive, and subtraction from the harder branch of an emission has a subleading effect on the value of \(C_1^{(\varsigma)}\), \Eq{plusdist_prsf_nlo} provides an approximate expression for the NLO distribution of \(C_1^{(\varsigma)}\) for any \PIRANHA{}-groomed jet.
    %
    Though the value of the effective grooming parameter may depend on the configuration of the particles for a more generic \PIRANHA{} groomer -- i.e. though in principle we must replace \(\zcut\to \zcut(z,\theta)\) -- we may use the soft and collinear limits to justify replacing \(\zcut(z,\theta)\) by \(\zcut(0,0)\) in \Eq{plusdist_prsf_nlo} for the applications considered in this thesis.
    %
    We can do even better by subtracting from both the hard and the soft branches, as we discuss in \Sec{pira-resummed} when performing calculations of \glspl{gecf} for jets groomed with \glslink{rs}{Recursive Subtraction (P-RS)}.
}

@@

    Argue that we can use \Eq{multiple-emissions} to realize a well-defined, continuous probability distribution for the constant-subtraction-groomed \gls{gecf} which includes the effects of multiple emissions.
\end{exercise}

@@

The EMD between a groomed jet in the presence and absence of low-energy pollution also provides an observable-independent tool for quantifying the sensitivity of jet grooming to pollution from \glspl{soft-distortion} and \gls{additive-contamination}.
%
Since the EMD furnishes a metric on the space of events, two events are separated by zero EMD iff they have identical energy flow, and two jets separated by zero EMD will yield the same value for every infra-red/collinear (IRC) safe observable (see \href{https://arxiv.org/pdf/2004.04159.pdf#page=11\&zoom=100,0,650}{Lemma 1} of \Reff{Komiske:2020qhg}).
%
Further, the EMD between two events also \textit{directly} bounds the differences in a large class of IRC-safe observable quantities (Lipschitz continuous observables) between the events~\cite{Komiske:2019fks}.
%
For example, the EMD between a parton-level jet and the same jet after \gls{hadronization} (the parton-hadron EMD) directly bounds the difference between a large class of parton- and hadron-level IRC-safe observables.
%
The parton-hadron EMD is therefore a powerful, observable independent tool for characterizing the response of a jet to \glspl{soft-distortion} from \gls{hadronization}.
%
We use the EMD in our analysis of grooming sensitivity to several sources of low-energy pollution in \Secs{soft_distortion}{add_contam}.

@@

    \item
        \glslink{angular-discontinuity}{\vocab{Angular Continuity}} is the requirement that a groomer is invariant under infinitesimally small angular changes in the directions of jet constituents.
\end{itemize}
\sam{Maybe this belongs earlier}

@@

The EMD itself also furnishes an example of an optimal transport problem, allowing us to immediately introduce a simple, generic \PIRANHA{} groomer.
%
To begin, let us model the radiation that we would like to remove from our jet as a distribution of energy \(\rhoC\) in a portion of the rapidity-azimuth (\(\eta\)-\(\phi\)) plane with area \(A\).
%
We use \(\rho\) to indicate the mean energy density and \(\mathcal{C}\) to indicate a distribution modeling the shape of the contaminating radiation, normalized such that
\begin{equation}
    \rho\int \dd y\,\dd\phi\,\mathcal{C}(y,\phi) = \rho A = E
\end{equation}
gives the total energy of the \gls{additive-contamination} we would like to remove.
%
Again echoing \Reff{Komiske:2020qhg}, we may phrase the subtraction of the energy distribution \(\rhoC\) from an event \(\mathcal{E}\) as the solution to the optimal transport problem
\begin{align}
    \label{eq:simple_piranha}
    \mathcal{E}_g[\mathcal E_0] = \underset{\mathcal{E}'}{\arg \min}\ {\rm EMD}_{\beta, R}
    \left(\mathcal{E}_0, \mathcal{E}' + \rhoC\right)
    .
\end{align}
%
It was shown in \href{https://arxiv.org/pdf/2004.04159.pdf#page=33\&zoom=100,0,0}{Lemma 3} of \Reff{Komiske:2020qhg} that \Eq{simple_piranha} defines a subtractive algorithm that merely decreases the energy of jet constituents, and does not add new particles to an event.
%
In the piranha analogy, \(\rhoC\) roughly describes the distribution of our piranhas, with a ``piranha energy density'' \(\rho\), while the minimization problem of \Eq{simple_piranha} provides them with the transport plans that guide them to subtract from the \(p_T\) of jet constituents.
%
Furthermore, the form of \(\mathcal{E}_g[\mathcal E_0]\) in \Eq{simple_piranha} is manifestly continuous due to its dependence on the EMD metric.%
\footnote{
    While \(\mathcal{E}_g[\mathcal E_0]\) in \Eq{simple_piranha} is continuous, manifest continuity due to dependence on the EMD should be treated skeptically.
    %
    For example, the ``grooming procedure'' associated with  \Eq{simple_piranha} may not be continuous if the \(\arg \min\) operation is restricted to a non-convex sub-space of the full space of events.
    %
    Analogously, \(
        f(x)
        =
        \underset{y}{\arg \min}\ ||x, y+1||
        =
        x-1
    \)
    is continuous as a map from \(\mathbb R\) to \(\mathbb R\) with the standard topology.
    %
    However, \(
        g(x)
        =
        \underset{y \in \{0, 1\}}{\arg \min}\ ||x, y+1||,
    \)
    which differs only by the allowed values of \(y\) in the \(\arg \min\), is not.
}

>>>>

The principle of \gls{lphd} (discussed in \Sec{phd}) suggests that small changes in hadronic energy flow should be associated with small changes in the associated partonic degrees of freedom, and vice versa.
%
Therefore, we will hope to achieve a way of isolating high-energy, partonic information in a \textit{continuous} way:
%
small changes in the flow of energy within a jet should not drastically affect what we infer about partonic degrees of freedom.
%
Conversely, a jet grooming algorithm is \textit{discontinuous} if small changes in the properties of jet constituents, such as their energies or angular positions, can lead to extensively different groomed results.

@@

    Therefore, the cumulative distribution at LL can be written
    \begin{equation}
    \begin{aligned}
        \label{eq:LL_mass_sd}
        \Sigma^{\rm(LL)}_{m^2}(m^2)
        =
        \sum_{n = 0}^{\infty}\frac{1}{n!}
        \Bigg(
        \frac{\alpha_s}{\pi}
        \int_0^R \frac{\dd\theta}{\theta}
        \int_0^{\frac{1}{2}} \dd z\,[\bar{p}_i(z)]_+^{[1/2]}
        \Bigg[
            \Theta\left(
                \frac{m^2}{Q^2} > z\theta^2
            \right)
            &
            \Theta\left(z > \zcut\right)
            \\
            \notag
            +
            &\,\,\Theta\left(z < \zcut\right)
        \Bigg]
        \Bigg)^n
        \,.
    \end{aligned}
    \end{equation}
    Using the properties of \glspl{plus-fn} as in \Sec{ll-substructure-diagrams}, we can conclude that
    \begin{equation}
    \begin{aligned}
        \Sigma^{\rm(LL)}_{m^2}(m^2)
        =
        \sum_{n = 0}^{\infty}\frac{1}{n!}
        \Bigg(
        \frac{\alpha_s}{\pi}
        \int_0^R \frac{\dd\theta}{\theta}
        \int_0^{\frac{1}{2}} \dd z\,\bar{p}_i(z)
        \Bigg[
            &\Theta\left(
                \frac{m^2}{Q^2} > z\theta^2
            \right)
            \Theta\left(z > \zcut\right)
            \\
            \notag
            &+
            \,\,\Theta\left(z < \zcut\right) - 1
        \Bigg]
        \Bigg)^n
        \,.
    \end{aligned}
    \end{equation}
    Finally, using that \(1 - \Theta(a) = \Theta(-a)\), we can use \(\Theta(z < \zcut) - 1 = -\Theta(z > \zcut)\) to notice that the integrand contains
    \begin{align}
        \le(
        \Theta\left(
            \frac{m^2}{Q^2} > z\theta^2
        \right)
        - 1
        \ri)
        \times
        \Theta\left(z > \zcut\right)
        =
        \Theta\left(
            z\theta^2 > \frac{m^2}{Q^2}
        \right)
        \times
        \Theta\left(z > \zcut\right)
        \,.
    \end{align}

@@

\remark{cc-discontinuity}{
    The discontinuity of the \gls{constant-cutoff} grooming algorithm manifests in LL calculations as a \emph{kinds in the radiators} of substructure observables, and associated discontinuities of LL probability distributions (e.g. as in \Eq{ll-cc-mass})
    %
    Unfortunately, the kink in the radiator presents a complication when attempting to achieve computations beyond LL.
    %
    In particular, the formula of \Eq{resummed-multiple-emissions} for the inclusion of multiple-emissions effects on cumulative distributions of additive observables depends on the derivative of the radiator.
    %
    We repeat it here for convenience:
    \begin{align}
        \Sigma_X(x)
        &=
        \frac{\exp\le(-R_X(x) - \gamma_E R'_X(x)\ri)}{\Gamma\le(1 + R'_X(x)\ri)}
        \,,
    \end{align}
    where \(R'\) denotes the negative logarithmic derivative of \(R\).
    %
    Therefore, probability distributions which leverage \Eq{resummed-multiple-emissions} depend on the \textit{second derivative} of the radiator.
    %
    Though the radiator turns out to have a well defined derivative, e.g. in \Eq{ll-cc-mass}, it is not second-differentiable, and it is more difficult to account for multiple-emissions effects in \glslink{hard-cutoff-groomer}{hard-cutoff-groomed} jet observables than in ungroomed jet observables.
}

@@

\gls{soft-drop} is a powerful tool which improves upon the strengths of the constant cutoff algorithm.
%
Like the \gls{constant-cutoff} algorithm, \gls{soft-drop} is a tree-based, angular-ordered \gls{declustering} algorithm,%
\footnote{
    As a reminder this means that \gls{soft-drop} acts on a jet with characteristic radius \(R_0\) that is first clustered using an arbitrary algorithm, such as the anti-\(k_t\) algorithm, and then re-clustered with the angular-ordered Cambridge-Aachen (C/A) algorithm.
    %
    As discussed in \Sec{reclustering}, this produces a binary tree of clustered particles within the jet, such that particles deeper in the tree tend to be closer in angle.
}
%
and is therefore computationally efficient and amenable to resummed calculations.
%
However, Soft Drop only needs to loop over a subset of the branches of the jet tree, unlike the \gls{constant-cutoff} algorithm, which loops over every branch.
%
Since common sources of contaminating radiation in jets tend to be wide angle, as touched upon in \App{pollution-models}, \gls{soft-drop} can be geared to more effectively remove low-energy pollution from collider events while retaining any low-energy information associated with the process of jet formation and the hard scattering.

@@

We would also like to understand the analytic behavior of the substructure of jets groomed with Soft Drop.
%
Luckily, the \glspl{substructure-diagram} we developed for use with the Constant Cutoff grooming algorithm can be quickly extended for use with Soft Drop.
%
Let us start with the most similar examples, for which \(\betasd=0\).

@@

We begin with a toy example of a \PIRANHA{} groomer in \Sec{constant-cutoff} -- the constant subtraction algorithm, similar to the constant cutoff algorithm of \Sec{constant-cutoff} -- before introducing the three concrete implementations of \PIRANHA{} discussed in this thesis:%
\footnote{
    Code for these three \PIRANHA{} groomers is available on GitHub \cite{piranhagithub}.
    %
    We delegate discussions of \PIRANHA{} grooming in EMD mode to \App{grooming_in_emd_mode}.
}
\begin{itemize}
    \item
        \vocab{\gls{rs}} is a family of nearly-continuous, computationally efficient, tree-based algorithms introduced in \Reff{Alipour-Fard:2023prj}.
    %
    We develop the concrete implementation of \vocab{\glslink{rsf}{Recursive Subtraction with a Fraction \(f_\text{soft}\) (P-RSF\(_{f}\))}} in \Sec{recursive-subtraction}.
    %
    We discuss how P-RSF\(_f\) is discontinuous in suppressed regions of parameter space, and focus on \PRSF{1/2}, which is the only fully soft-continuous version of P-RSF\(_f\);

@@

Before we give a more detailed account of \PIRANHA{} grooming, however, let us start with some guiding philosophy.
%
Given a measured collider event \(\mathcal{E}_0\) and a model \(\rhoC\) for the energy density of low energy pollution, we roughly want to subtract out the contamination to obtain a groomed collider event \(\mathcal{E}_g\);
%
na\"ively, we wish to write, formally,
\begin{align}
    \label{eq:naive-subtraction}
    \mathcal{E}_g[\mathcal{E}_0]
    =
    \mathcal{E}_0
    -
    \rhoC
    \,.
\end{align}
However, strict point-wise subtraction (i.e. \(\mathcal{E}_g(\hat{n}) = \mathcal{E}_0(\hat{n}) -\rhoC(\hat{n})\)) is not available to us if we want \(\mathcal{E}_g\) to be a well-defined energy flow:
%
subtraction can produce negative numbers, and negative energies are not allowed in the space of energy flows.
%
However, based on the motivation above, we want a grooming algorithm that behaves \emph{roughly} through subtraction.
%
This motivation will push us to define the constant subtraction and Recursive Subtraction algorithms of \Secs{constant-subtraction}{recursive-subtraction}.

In \Sec{emd}, however, we will introduce technology that will allow us to go further and take \Eq{naive-subtraction} nearly literally.
%
In particular, we will introduce the EMD:
%
a \emph{metric} that enables us to generalize the notion of subtraction.
%
Replacing \(\mathcal{E}\) by \(E\) and \(\rhoC\) by \(c\) for subtraction on the real line, we can write \(E_0 - c = \text{argmin}_{E_g'}d(E_g' + c, E_0)\), where \(d(\cdot,\cdot)\) is the standard metric on the real line.
%
This complicated way of writing regular subtraction of real numbers gives us the correct intuition for how to proceed with the EMD:
%
given an ungroomed event with energy density \(\mathcal{E}_{0}\) in the pseudorapidity-azimuth angle plane and any ansatz \(\rhoC\) for the energy density of low-energy pollution, one particularly succinct \PIRANHA{} algorithm is
%
\begin{align}
    \mathcal{E}_g\left[\mathcal{E}_0\right]
    =
    \underset{\mathcal{E}'}{\arg \min}\ {\rm EMD}_{\beta, R}
    \left(\mathcal{E}_0, \mathcal{E}' + \rhoC\right)
    \label{eq:simple_piranha_intro}
    \,,
\end{align}
where explicit dependent on the EMD metric makes the continuity of \(\mathcal{E}_g\left[\mathcal{E}_0\right]\) manifest.
%
Because the EMD is an optimal transport algorithm, \PIRANHA{} groomers may be roughly imagined as optimally transporting hungry piranhas to eat up delicious low-energy pollution.

\gls{apollonius} is a direct, continuous application of \Eq{simple_piranha_intro} with \(\beta = 1\), and for which \(\rhoC\) is uniform in the pseudorapidity-azimuth plane to approximate common sources of \gls{additive-contamination} such as \gls{pileup} and \gls{ue} \cite{Soyez:2018opl,Monk:2018clo,Sjostrand:1987su,Sjostrand:2014zea,Dasgupta:2007wa,Kirchgaesser:2020poq,Moraes:2007rq,CDF:2015txs,Larkoski:2021hee,Baron:2020xoi,Marzani:2017kqd}.
%
\gls{ivs} approximates \gls{apollonius} and preserves its continuity.
%
Finally, the \glslink{rsf}{Recursive Subtraction with a Fraction 1/2 (\PRSF{1/2})} algorithm we introduce in \Sec{recursive-subtraction} furnishes, to our knowledge, the first example of a soft continuous tree-based jet grooming algorithm;
%
the soft discontinuities of other P-RS algorithms occur only in highly suppressed regions of phase space.

This section is devoted to an introduction of the \PIRANHA{} paradigm.
%
Readers looking for a more detailed discussion of continuity may want to skip to \Sec{discontinuity}, which examines the soft discontinuities present in hard-cutoff groomers and \gls{rsf}, the soft continuity of \PRSF{1/2}, and angular discontinuity in tree-based groomers.

\remark{}{
    Our presentation, beginning with the constant subtraction and P-RS algorithms \emph{before} introducing \gls{apollonius}, is chronologically backwards and chosen with the goal of pedagogical clarity.
    %
    \gls{apollonius} and \gls{ivs} were both introduced in \Reff{Komiske:2020qhg}, and myself and my collaborators introduced P-RS and the \PIRANHA{} paradigm in the follow-up work of \Reff{Alipour-Fard:2023prj}.
    %
    The constant subtraction algorithm that we begin with now is not really of formal interest because it is collinear unsafe;
    %
    however, it exposits the basics of subtractive algorithms without too many algorithmic steps, and is conceptually similar to the constant cutoff algorithm we introduced in \Sec{constant-cutoff}.
    %
    For this reason, we now use the constant subtraction algorithm to introduce the basics of our calculations of \PIRANHA{}-groomed substructure.
}

@@

The constant subtraction algorithm is not quite a \PIRANHA{} algorithm -- it is not even collinear safe;
%
nonetheless, additive substructure observables computed on constant-subtraction groomed jets \textit{are} collinear safe, and the computation of additive observables in the simple context of this section will give us a sneak peak of computations of jet substructure in the Recursive Subtraction algorithms we explore in \Sec{recursive-subtraction}.

@@

Though the constant subtraction algorithm is not technically a \PIRANHA{} algorithm, we are starting to see why the name \PIRANHA{} is suitable;
%
not only does this algorithm annihilate the soft particles associated with \glslink{pileup}{pileup} and infra-red radiation, but it also can be imagined as putting a fish on each softer branch with a hunger for energy (fraction) given by \zcut, and which will subtract from its assigned branch until either it eats its fill or it completely annihilates its meal.

@@

Up to terms that are power-suppressed in \(C\), \(\zcut\), or both, the fixed-order distribution for the groomed observable away from \(C = 0\) is
%
\begin{subequations}
\label{eq:piranha_nlo}
\begin{align}
    \rho^{\rm(NLO)}_{i,~\varsigma}(C>0)
    &=
    \frac{2\alpha_s C_{R_i}}{\varsigma~\pi}
    \frac{1}{C}
    \left(
        2\tanh^{-1}\left(1  -2 \tilde C - 2\zcut\right)
        + B_i
        +
        \mathcal{O}\left(C,\,\zcut\right)
    \right)
    \\
    &=
    \frac{2\alpha_s C_{R_i}}{\varsigma~\pi}
    \frac{1}{C}
    \left(
        -\log\left(\tilde C + \zcut\right)
        + B_i
        +
        \mathcal{O}\left(C,\,\zcut\right)
    \right)
\end{align}
\end{subequations}
%
where \(\tilde C = C \times (1-\zcut)^2\), \(C_{R_i}\) is the quadratic Casimir of the color representation of the initiating parton, and we have defined the factor \(B_i\) to capture behavior from non-singular components of the splitting functions:
%
\(B_q = -3/4\) for quark initiated jets and \(B_g = -11/12 + N_f/(6 C_A)\) for gluon initiated jets.
\end{example}

@@

\noindent
The plus-function regularized expression of \Eq{plusdist_prsf_nlo} may be derived by replacing \(\overline{p}(z)\) in \Eq{NLO_dist_setup} with its plus-function analog.
%
Alternatively, one can recognize that the full distribution for \(C_1^{(\varsigma)}\) must integrate to one even at \glslink{accuracy}{NLO}, and that additional contributions to the \glslink{accuracy}{NLO} distribution in \Eq{prsf_NLO} can only come from the regions of phase space where \(C_1^{(\varsigma)}=0\);
%
this \gls{sum-rule} is enough to uniquely specify the result of \Eq{plusdist_prsf_lo} at \glslink{accuracy}{NLO}.
\end{exercise}

\remark{}{
    Since \PIRANHA{} algorithms are subtractive, and subtraction from the harder branch of an emission has a subleading effect on the value of \(C_1^{(\varsigma)}\), \Eq{plusdist_prsf_nlo} provides an approximate expression for the NLO distribution of \(C_1^{(\varsigma)}\) for any \PIRANHA{}-groomed jet.
    %
    Though the value of the effective grooming parameter may depend on the configuration of the particles for a more generic \PIRANHA{} groomer -- i.e. though in principle we must replace \(\zcut\to \zcut(z,\theta)\) -- we may use the soft and collinear limits to justify replacing \(\zcut(z,\theta)\) by \(\zcut(0,0)\) in \Eq{plusdist_prsf_nlo} for the applications considered in this thesis.
    %
    We can do even better by subtracting from both the hard and the soft branches, as we discuss in \Sec{pira-resummed} when performing calculations of \glspl{gecf} for jets groomed with \glslink{rs}{Recursive Subtraction (P-RS)}.
}

@@

    Argue that we can use \Eq{resummed-multiple-emissions} to realize a well-defined, continuous probability distribution for the constant-subtraction-groomed \gls{gecf} which includes the effects of multiple emissions.
\end{exercise}

@@

The EMD between a groomed jet in the presence and absence of low-energy pollution also provides an observable-independent tool for quantifying the sensitivity of jet grooming to pollution from \glspl{soft-distortion} and \gls{additive-contamination}.
%
Since the EMD furnishes a metric on the space of events, two events are separated by zero EMD iff they have identical energy flow, and two jets separated by zero EMD will yield the same value for every infra-red/collinear (IRC) safe observable (see \href{https://arxiv.org/pdf/2004.04159.pdf#page=11\&zoom=100,0,650}{Lemma 1} of \Reff{Komiske:2020qhg}).
%
Further, the EMD between two events also \textit{directly} bounds the differences in a large class of IRC-safe observable quantities (Lipschitz continuous observables) between the events~\cite{Komiske:2019fks}.
%
For example, the EMD between a parton-level jet and the same jet after \gls{hadronization} (the parton-hadron EMD) directly bounds the difference between a large class of parton- and hadron-level IRC-safe observables.
%
The parton-hadron EMD is therefore a powerful, observable independent tool for characterizing the response of a jet to \glspl{soft-distortion} from \gls{hadronization}.
%
We use the EMD in our analysis of the sensitivity of grooming to several sources of low-energy pollution in Sections \ref{sec:pira-hadronization} through \ref{sec:pira-ue}.

@@

    \item
        \glslink{angular-discontinuity}{\vocab{Angular Continuity}} is the requirement that a groomer is invariant under infinitesimally small angular changes in the directions of jet constituents.
\end{itemize}

@@

The EMD itself also furnishes an example of an optimal transport problem, allowing us to immediately introduce a simple, generic \PIRANHA{} groomer.
%
To begin, let us model the radiation that we would like to remove from our jet as a distribution of energy \(\rhoC\) in a portion of the rapidity-azimuth (\(\eta\)-\(\phi\)) plane with area \(A\).
%
We use \(\rho\) to indicate the mean energy density and \(\mathcal{C}\) to indicate a distribution modeling the shape of the contaminating radiation, normalized such that
\begin{equation}
    \rho\int \dd y\,\dd\phi\,\mathcal{C}(y,\phi) = \rho A = E
\end{equation}
gives the total energy of the \gls{additive-contamination} we would like to remove.
%
Again echoing \Reff{Komiske:2020qhg}, we may phrase the subtraction of the energy distribution \(\rhoC\) from an event \(\mathcal{E}\) as the solution to the optimal transport problem
\begin{align}
    \label{eq:simple_piranha}
    \mathcal{E}_g[\mathcal E_0] = \underset{\mathcal{E}'}{\arg \min}\ {\rm EMD}_{\beta, R}
    \left(\mathcal{E}_0, \mathcal{E}' + \rhoC\right)
    .
\end{align}
%
It was shown in \href{https://arxiv.org/pdf/2004.04159.pdf#page=33\&zoom=100,0,0}{Lemma 3} of \Reff{Komiske:2020qhg} that \Eq{simple_piranha} defines a subtractive algorithm that merely decreases the energy of jet constituents, and does not add new particles to an event.
%
In the piranha analogy, \(\rhoC\) roughly describes the distribution of our piranhas, with a ``piranha energy density'' \(\rho\), while the minimization problem of \Eq{simple_piranha} provides them with the transport plans that guide them to subtract from the \(p_T\) of jet constituents.
%
Furthermore, the form of \(\mathcal{E}_g[\mathcal E_0]\) in \Eq{simple_piranha} is manifestly continuous due to its dependence on the EMD metric.

\remark{care-with-continuity}{
    While \(\mathcal{E}_g[\mathcal E_0]\) in \Eq{simple_piranha} is continuous, manifest continuity due to dependence on the EMD should be treated skeptically.
    %
    For example, the ``grooming procedure'' associated with  \Eq{simple_piranha} may not be continuous if the \(\arg \min\) operation is restricted to a non-convex sub-space of the full space of events.
    %
    Analogously, \(
        f(x)
        =
        \underset{y}{\arg \min}\ ||x, y+1||
        =
        x-1
    \)
    is continuous as a map from \(\mathbb R\) to \(\mathbb R\) with the standard topology.
    %
    However, \(
        g(x)
        =
        \underset{y \in \{0, 1\}}{\arg \min}\ ||x, y+1||,
    \)
    which differs only by the allowed values of \(y\) in the \(\arg \min\), is not.
}
